What is an Ancestral Recombination Graph?

### 2022-03-10 paper.tex

[We should discuss the relationship between ARG nodes and SPRs
here. Cite \citep{deng2021distribution} too. Maybe we want to make
this a separate section, as it's a subtle and important point.]

>>>>

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a stocastic process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
which are central to many modern methods.

Under the coalescent with recombination (or SMC) each local
tree along the genome is transformed into the next by
a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the "subtree transfer" operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

[Discussion of SPRs in an EARG]

The effects of recombination as we move from left-to-right along
the genome in an unsimplified GARG are easy to interpret.
Consider the leftmost tree in Fig~\ref{fig-simplification}.
In the first tree the parent of B is F, and in the second
tree its parent is E. Since both parents are present as
unary nodes in the tree, we simply change one edge in
order to effect the transition.
A more complex transition occurs when we have a regraft operation
that occurs above the current root (i.e., the current oldest
non-unary node). For example, the transition between the
second and third trees corresponds to node D switching
from the path through I to H, and we then must add branches
corresponding to passing through L and meeting at M.

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is important to note that while extracting local
trees from an ARG is straightforward, the converse
process of reconstructing precisely the same ARG from
these trees is more problematic. If the internal
nodes are consistently labelled and unary nodes are
included in the local trees (e.g., Fig~\ref{fig-simplification})
then it is easy to exactly reconstruct the ARG.
Otherwise, the problem is much more difficult and it
may not be possible to uniquely reconstruct the original ARG.
Without labels for the internal nodes, we are forced to
to reason about which subtrees remain the same and which
differ due to the SPR separating adjacent trees in order to
assign node identity. Without the internal unary nodes marking
recombinations, we must infer their existence. In
general, there will not be a unique solution to this problem,
and so there can be many possible ARGs compatible with
a given list of
% want to say "standard phylogenetic" trees here or something, but
% and they don't have to be binary (just arity > 1), but thought
% this would be familiar to readers?
% Jere: Could just say leaf-labelled "marginal trees", "ancestral trees",
% "local trees", or something like that.
leaf-labelled binary trees.

[ Cite \citep{deng2021distribution} somewhere here.]
